---
output: 
  html_document: 
    keep_md: yes
    self_contained: no
---

## Simple Features

A standardized set of geometric shapes are the essence of vector data. 
The [sf](){:.rlib} puts sets of shapes in a tabular structure so that we can manipulate and analyze them.

```{r, handout = 0, message = FALSE}
library(sf)

lead <- read.csv('data/SYR_soil_PB.csv')
lead[['geometry']] <- st_sfc(
  st_point(),
  crs = 32618)
```

===

We read in a data frame called `lead` from an external CSV file, then create a new column called `geometry`. 
The `lead` data frame now has the "simple feature column", which `st_sfc` creates from a CRS and a geometry type 
(in this case a point geometry, `st_point()`). For now, each point is "EMPTY". The empty geometry is 
equivalent to a NA value. *Note*: you must call this column `geometry` so that other [sf](){:.rlib} functions
will recognize it as the simple feature column.
{:.notes}

```{r}
head(lead)
```

===

| geometry | description |
|----------+-------------|
| `st_point` | a single point |
| `st_linestring` | sequence of points connected by straight, non-self-intersecting line pieces |
| `st_polygon` | one [or more] sequences of points in a closed, non-self-intersecting ring [with holes] |
| `st_multi*` | sequence of `*`, either `point`, `linestring`, or `polygon` |

===

Each element of the simple feature column is a simple feature geometry, here
created from the "x" and "y" elements of a given feature.

Here we use the first entry in the existing `x` and `y` columns of the `lead` 
data frame to fill in the empty `geometry` slot in the first row of the data frame.
The `dim = 'XY'` argument specifies a two-dimensional point.
{:.notes}

```{r, handout = 0 }
lead[[1, 'geometry']] <- st_point(
  c(x = lead[[1, 'x']],
    y = lead[[1, 'y']]),
  dim = 'XY')
```

```{r}
head(lead)
```

===

The whole data frame must be cast to a simple feature object, which causes
functions like `print` and `plot` to use methods introduced by the `sf` library.

```{r, handout = 0 }
lead <- st_sf(lead)
```

===

For example, the `print` method automatically shows the CRS and truncates the
number of records displayed.

```{r}
lead
```

===

Naturally, there is a shortcut to creating an `sf` object from a data frame with
point coordinates. We use `st_as_sf()` with the `coords` argument to specify
which columns represent the x and y coordinates of each point.
The CRS must be specified with the argument `crs` via EPSG integer or proj4 string.

```{r, handout = 0 }
lead <- read.csv('data/SYR_soil_PB.csv')
lead <- st_as_sf(lead,
  coords = c('x', 'y'),
  crs = 32618)
```

```{r}
lead
```

===

Now that the data frame is an `sf` object, the data are easily displayed as a map.

Here, the points are colored by lead concentration (`ppm`).
{:.notes}

```{r, handout = 0 }
plot(lead['ppm'])
```

===

For [ggplot2]({:.rlib}) figures, use `geom_sf` to draw maps. In the `aes` mapping for feature collections, 
the "x" and "y" variables are automatically assigned to the "geometry" column, 
while other attributes can be assigned to visual elements like `fill` or `color`.

===

```{r, handout = 0}
library(ggplot2)

ggplot(data = lead,
       mapping = aes(color = ppm)) +
  geom_sf()
```

===

## Feature Collections

More complicated geometries are usually not stored in CSV files, but they are
usually still read as tabular data. We will see that the similarity of feature
collections to non-spatial tabular data goes quite far; the usual data
manipulations done on tabular data work just as well on `sf` objects. 
Here, we read the boundaries of the [Census block groups](https://www.census.gov/programs-surveys/geography/about/glossary.html#par_textimage_4)
in the city of Syracuse from a shapefile.
{:.notes}

```{r, handout = 0 }
blockgroups <- read_sf('data/bg_00')
```

===

The `geometry` column contains projected UTM coordinates of the polygon vertices.

```{r}
blockgroups
```

===

Also note the table dimensions show that there are 147 features in the collection.

```{r}
dim(blockgroups)
```

===

Simple feature collections are data frames.

```{r}
class(blockgroups)
```

===

The `blockgroups` object is a `data.frame`, but it also has the class attribute
of `sf`. This additional class extends the `data.frame` class in ways useful for
feature collection. For instance, the geometry column becomes "sticky" in most
table operations, like subsetting. This means that the geometry column is retained
even if it is not explicitly named.
{:.notes}

```{r}
blockgroups[1:5, 'BKG_KEY']
```

===

## Table Operations

We can plot the polygon features using [ggplot2](){:.rlib} or do common
data wrangling operations:
{:.notes}

- plot
- merge or join
- "split-apply-combine" or group-by and summarize

===

```{r}
ggplot(blockgroups,
       aes(fill = Shape_Area)) +
  geom_sf()
```

===

Merging with a regular data frame is done by normal merging on non-spatial
columns found in both tables.

```{r, handout = 0 }
census <- read.csv('data/SYR_census.csv')
census <- within(census, {
     BKG_KEY <- as.character(BKG_KEY)
})
```

Here, we read in a data frame called `census` with demographic characteristics
of the Syracuse block groups.
As usual, there's the difficulty that CSV files do not include metadata on data
types, which have to be set manually. We do this here by coercing the `BKG_KEY`
column to character format using `as.character()`.
{:.notes}

===

Merge tables on a unique identifier (our primary key is "BKG_KEY"), but let the
"sf" object come first or its special attributes get lost.

The `inner_join()` function from [dplyr](){:.rlib} joins two data frames on a
common key specified with the `by` argument, keeping only the rows from each data 
frame with matching keys in the other data frame and discarding the rest.
{:.notes}

```{r, handout = 0, message = FALSE, warning = FALSE}
library(dplyr)

census_blockgroups <- inner_join(
  blockgroups, census,
  by = c('BKG_KEY'))
```

```{r}
class(census_blockgroups)
```

===

The census data is now easily visualized as a map.

```{r, handout = 0}
ggplot(census_blockgroups,
       aes(fill = POP2000)) +
  geom_sf()
```

===

Feature collections also cooperate with the common "split-apply-combine" sequence of steps in data manipulation.

- *split* -- group the features by some factor
- *apply* -- apply a function that summarizes each subset, including their geometries
- *combine* -- return the results as columns of a new feature collection

===

```{r, handout = 0 }
census_tracts <- census_blockgroups %>%
  group_by(TRACT) %>%
  summarise(
    POP2000 = sum(POP2000),
    perc_hispa = sum(HISPANIC) / POP2000)
```

Here, we use the [dplyr](){:.rlib} function `group_by` to specify that we are doing an operation
on all block groups within a [Census tract](https://www.census.gov/programs-surveys/geography/about/glossary.html#par_textimage_13).
We can use `summarise()` to calculate as many summary statistics as we want, each one separated by a comma.
Here `POP2000` is the sum of all block group populations in each tract and `perc_hispa` is the sum of the Hispanic 
population divided by the total population (because now we have redefined `POP2000` to be the population total). 
The `summarise()` operation automatically combines the block group polygons from each tract!
{:.notes}

===

Read in the census tracts from a separate shapefile to confirm that the
boundaries were dissolved as expected.

```{r, handout = 0 }
tracts <- read_sf('data/ct_00')
ggplot(census_tracts,
       aes(fill = POP2000)) +
  geom_sf() +
  geom_sf(data = tracts,
          color = 'red', fill = NA)
```

===

By default, the sticky geometries are summarized with `st_union`. The
alternative `st_combine` does not dissolve internal boundaries. Check
`?summarise.sf` for more details.
