---
---

## Regression

Continuing the theme that vector data *is* tabular data, the natural
progression in statistical analysis is toward regression. Building a regression
model requires making good assumptions about relationships in your data:

- between *columns* as independent and dependent variables
- between *rows* as more-or-less independent observations

===

The following model assumes an association (in the linear least-squares sense),
between the Hispanic population and lead concentrations and assumes independence
of every census tract (i.e. row).

```{r, handout = 0}
ppm.lm <- lm(pred_ppm ~ perc_hispa,
  data = census_lead_tracts)
```

===

Is that model any good?

```{r, handout = 0}
census_lead_tracts <- census_lead_tracts %>%
  mutate(lm.resid = resid(ppm.lm))
plot(census_lead_tracts['lm.resid'])
```

===

We can see that areas with negative and positive residuals are not randomly distributed.
They tend to cluster together in space: there is
autocorrelation. It is tempting to ask for a semivariogram plot of the
residuals, but that requires a precise definition of the distance between
polygons. A favored alternative for quantifying autoregression in non-point
feature collections is Moran's I. This analog to Pearson's correlation
coefficient quantifies autocorrelation rather than cross-correlation.
{:.notes}

Moran's I is the correlation between all pairs of features, weighted to
emphasize features that are close together. It does not dodge the problem of
distance weighting, but provides default assumptions for how to do it.

```{r, handout = 0, message = FALSE}
library(sp)
library(spdep)
library(spatialreg)

tracts <- as(
  st_geometry(census_tracts), 'Spatial')
tracts_nb <- poly2nb(tracts)
```

The function `poly2nb()` from [spdep](){:.rlib} generates a neighbor list from
a set of polygons. This object is a list of each polygon's neighbors that share
a border with it (in matrix form, this would be called an adjacency matrix).
Unfortunately, [spdep](){:.rlib} was created before [sf](){:.rlib}, so the two 
packages aren't compatible. That's why we first need to convert the geometry of `census_tracts` 
to an [sp](){:.rlib} object using `as(st_geometry(...), 'Spatial')`.
{:.notes}

===

The `neighbors` variable is the network of features sharing a boundary point.

```{r, handout = 0}
plot(census_lead_tracts['lm.resid'],
     reset = FALSE)
plot.nb(tracts_nb, coordinates(tracts),
        add = TRUE)
```

`plot.nb()` plots a network graph, given a `nb` object and a matrix of coordinates as 
arguments. We extract a coordinate matrix from the `Spatial` object `tracts` using
`coordinates()`.
{:.notes}

===

Add weights to the neighbor links.

```{r, handout = 0}
tracts_weight <- nb2listw(tracts_nb)
```

By default, `nb2listw()` weights each neighbor of a polygon equally, although other
weighting options are available.
{:.notes}

===

Visualize correlation between the residuals and the weighted average
of their neighbors with `moran.plot` from the
[spdep](){:.rlib}. The positive trend line is consistent with the
earlier observation that features in close proximity have similar
residuals.

```{r, handout = 0}
moran.plot(
  census_lead_tracts[['lm.resid']],
  tracts_weight,
  labels = census_lead_tracts[['TRACT']],
  pch = 19)
```

The first argument to `moran.plot` is the vector of data values (in this case the residuals),
and the second argument is the weighted neighbor list. The `labels` argument in this
case is the vector of tract codes. By default `moran.plot` will flag and label values
that have high influence on the relationship.
{:.notes}

===

There are many ways to use geospatial information about tracts to impose
assumptions about non-independence between observations in the regression. One
approach is a Spatial Auto-Regressive (SAR) model, which regresses each value against
the weighted average of neighbors.

```{r, include = FALSE}
set.seed(20190327)
```
```{r, handout = 0}
ppm.sarlm <- lagsarlm(
  pred_ppm ~ perc_hispa,
  data = census_lead_tracts,
  tracts_weight,
  tol.solve = 1.0e-30)
```

Here, `lagsarlm()` from [spatialreg](){:.rlib} uses the same formula as the `lm()` 
we did above, but we also need to supply the weighted neighbor list.
The `tol.solve` argument is needed for the numerical method the model uses.
{:.notes}

===

The Moran's I plot of residuals now shows less correlation; which means the SAR
model's assumption about spatial autocorrelation (i.e. between table rows) makes
the rest of the model more plausible.

```{r, handout = 0}
moran.plot(
  resid(ppm.sarlm),
  tracts_weight,
  labels = census_lead_tracts[['TRACT']],
  pch = 19)
```

===

Feeling more confident in the model, we can now take a look at the regression
coefficients and overall model fit.

```{r}
summary(ppm.sarlm)
```


