---
---
    
## The Semivariogram

A semivariogram quantifies the effect of distance on the correlation between
values from different locations. *On average*, measurements of the same variable
at two nearby locations are more similar (lower variance) than when those
locations are distant.
{:.notes}

The [gstat](){:.rlib} library, a geospatial analogue to the [stats](){:.rlib}
library, provides variogram estimation, among several additional tools.

```{r, handout = 0}
library(gstat)
lead_xy <- read.csv('data/SYR_soil_PB.csv')
```

===

The empirical semivariogram shown below is a windowed average of the squared
difference in lead concentrations between sample points.
As expected, the difference increases as distance between points increases and
eventually flattens out at large distances.

```{r, handout = 0}
v_ppm <- variogram(
  ppm ~ 1,
  locations = ~ x + y,
  data = lead_xy)
plot(v_ppm)
```

Here we use the formula `ppm ~ 1` to specify that lead concentration `ppm` is
the dependent variable, and we are not fitting any trend (1 means intercept only).
We pass another formula to the `locations` argument, `locations = ~ x + y`,
to specify the x and y coordinate of each location.
{:.notes}

===

Fitting a model semivariogram tidies up the information about autocorrelation
in the data, so we can use it for interpolation.

```{r, handout = 0}
v_ppm_fit <- fit.variogram(
  v_ppm,
  model = vgm(1, "Sph", 900, 1))
plot(v_ppm, v_ppm_fit)
```

We use `fit.variogram()` to fit a model to our empirical variogram `v_ppm`. The type of 
model is specified with the argument `model = vgm(...)`.
Describing the parameters of `vgm()` used to fit the model variogram is outside the scope of this
lesson; for more information check out [gstat documentation](https://cran.r-project.org/web/packages/gstat/index.html) 
and the references cited there.
{:.notes}

===

## Kriging

The modeled semivariogram acts as a parameter when performing Gaussian process regression, commonly known as kriging, to
interpolate predicted values for locations near our observed data. 
The steps to perform kriging with the [gstat](){:.rlib} library are:

1. Generate a modeled semivariogram
1. Generate new locations for "predicted" values
1. Call `krige` with the data, locations for prediction, and semivariogram

===

Generate a low resolution (for demonstration) grid of points overlaying the
bounding box for the lead data and trim it down to the polygons of interest.
Remember, that goal is aggregating lead concentrations within each census tract.

```{r, handout = 0}
pred_ppm <- st_make_grid(
  lead, cellsize = 400,
  what = 'centers')
idx <- unlist(
  st_intersects(census_tracts, pred_ppm))
pred_ppm <- pred_ppm[idx]
```

Here, `st_make_grid()` creates a grid of points `pred_ppm` at the center of 400-meter squares over the
extent of the `lead` geometry. We can specify `cellsize = 400` because the CRS of `lead` is
in units of meters, and `what = 'centers'` means we want the points at the center of each
grid cell rather than polygons representing the cells. Next, we find only the points contained
within census tract polygons using `st_intersects(census_tracts, pred_ppm)`, then `unlist()`
to return a vector of indices `idx` for the points we want. Finally we subset `pred_ppm` by `idx`.
{:.notes}

===

Map the result to verify that we have only points contained in our census tracts.

```{r, handout = 0}
ggplot(census_tracts,
       aes(fill = POP2000)) +
  geom_sf() +
  geom_sf(data = pred_ppm,
          color = 'red', fill = NA)
```

===

Almost done ...

1. Generate a modeled semivariogram
1. Generate new locations for "predicted" values
1. Call `krige` with the data, locations, and semivariogram

===

The first argument of `krige` specifies the model for the means, which is constant according to our 
formula for lead concentrations (again, `~ 1` means no trend). 
The observed ppm values are in the `locations` data.frame along with the point geometry. 
The result is a data frame with predictions for lead concentrations at the points in `newdata`.

```{r, handout = 0, results = 'hide'}
pred_ppm <- krige(
  formula = ppm ~ 1,
  locations = lead,
  newdata = pred_ppm,
  model = v_ppm_fit)
```

===

And the same commands that joined lead concentrations to census tracts apply to
joining the predicted concentrations in too.

```{r, handout = 0, message = FALSE}
pred_ppm_tracts <-
  pred_ppm %>%
  st_join(census_tracts) %>%
  st_drop_geometry() %>%
  group_by(TRACT) %>%
  summarise(pred_ppm = mean(var1.pred))
census_lead_tracts <- 
  census_lead_tracts %>%
  inner_join(pred_ppm_tracts)
```

===

The predictions should be, and are, close to the original averages with
deviations that correct for autocorrelation.

```{r, handout = 0}
ggplot(census_lead_tracts,
       aes(x = pred_ppm, y = avg_ppm)) +
  geom_point()
```

===

The effect of paying attention to autocorrelation is subtle, but it is noticeable and had the expected effect in tract 5800. The pred_ppm value is a little higher than the average.

```{r}
census_lead_tracts[52,]
```
